---
title: "Main1.withoutCompetition.simulation"
author: "ENAT"
date: "2025-06-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

library(here)
rm(list=ls())


###########################
# 0 # file/folder paths and package loading
# you may need to change them to the paths on your computer before testing the code.
dir.create(here("data", "processed", "simulatedData"))
wdmm = here("data", "processed", "simulatedData")# a path to the folder you want to save the results.

wdinput=paste0(wdmm,"/SpeciesPool") # define subfolder which has regional pool
code.wd= here("scripts", "iCAMP", "simulation")
save.wd=paste0(wdmm,"/NoComp") 
if(!dir.exists(save.wd)){dir.create(save.wd)}


source(paste0(code.wd,"/tools.r"))
source(paste0(code.wd, "/simulation.functions.r"))
library(ape)
library(vegan)
library(bigmemory)
library(iCAMP)
library(magrittr)
library(dplyr)

```

```{r}

#load tree and trait(OPEN) data

tp.file= here("results", "output_data", "iCAMP_simulation", "JS.tree.pd.rda")
tp.in =  lazyopen(tp.file)
tree=tp.in$tree
pd=tp.in$pd
tree;max(tree$edge.length)
maxpd=max(pd)
op.file= here("results", "output_data", "iCAMP_simulation", "JS.opens.rda")
ops= lazyopen(op.file)
nworker=8

#YXIN_0h_metadata <- read.delim(here("data", "processed", "YXIN_QIIME_output", 
#                                    "XYIN_16S_515F_926R_07092023_Q16712_XYIN@metadata_formatted.txt")) %>%
#  as.data.frame() %>%
#  filter(time == "0h") %>%
# mutate(source.sampleID = paste0(sludge_source, "_", SampleID))

# setting gobal parameters
sampname = c(paste0("Almere",1:5),paste0("Bath",1:5))
Nk=20000 # individual number in each community
J=Nk*1000
theta=5000
spnumi=100

ds = 0.2
d.cut=(maxpd-ds)/2
op = ops$BM$openv
bin.size = 12
prefix=paste0("BMb",bin.size,"d",sub("\\.","",ds))

# simulate species abundance disitribution
MZSM.meta=sads::rmzsm(n=length(op),J=Nk*1000,theta=theta)
save(MZSM.meta,file=paste0(save.wd, "/MZSM.meta.J",J/1000/1000,"M.theta",theta,".rda"))

  
```

``` {r}

# 2 # Scenario 1: selection & weak selection

FL=0.05
FH=0.95
sig2.L=0.015
sig2.H=0.015


comm=sim_select(op,Nk,sampname,code.wd,FL=FL,FH=FH,sig2.L=sig2.L,sig2.H=sig2.H)
dim(comm)
prefixi=paste0(prefix,".S1.Sel")
ABP=ABPcount(comm = comm, sel.spname = colnames(comm))
save.sim(prefixi,comm,pd,tree,save.wd,ABP)
  
```

```{r}

# 3 # Scenario 2: extreme dispersal
#######################################
m1=0.01
m2=0.99

comm=dispersal.ZSM.sloan(op,Nk,sampname,spnumi=100,meta.ab=MZSM.meta,m1=m1,m2=m2,distinct=TRUE,fix.rich = TRUE)
dim(comm)
prefixi=paste0(prefix,".S2.Disp")
ABP=ABPcount(comm = comm, disp.spname = colnames(comm))
save.sim(prefixi,comm,pd,tree,save.wd,ABP)
  
```

``` {r}

# 4 # Scenario 3: drift
  #######################################
  # Sloan's dispersal model with MZSM metacommunity
  #######################################
  m1=0.5
  m2=0.01
  sig2.W=4
  
  comm=drift.ZSM.sloan(op,Nk,sampname,spnumi=100,meta.ab=MZSM.meta,
                       m1=m1,m2=m2,FH=0.95,FL=0.05,sig2.W=sig2.W)
  dim(comm)
  # test #
  prefixi=paste0(prefix,".S3.Drift")
  ABP=ABPcount(comm = comm, drift.spname = colnames(comm))
  save.sim(prefixi,comm,pd,tree,save.wd, ABP)

```

``` {r}

# 5 # Scenario 4-6: some species under extreme dispersal, some under drift
  
  s46.sim<-function(a.disp,a.drift,prefixi,...)
  {
    id.disp=sample(1:length(op),length(op)*a.disp)
    meta.disp=MZSM.meta[id.disp]
    meta.drift=MZSM.meta[-id.disp]
    op.disp=rep(1,length(id.disp));names(op.disp)=paste0("disp",1:length(id.disp))
    op.drift=rep(1,length(op)-length(id.disp));names(op.drift)=paste0("drift",1:length(op.drift))
    
    Nk.disp=Nk*a.disp
    Nk.drift=Nk*a.drift
    spnumi.disp=spnumi*a.disp
    spnumi.drift=spnumi*a.drift
    
    m1=0.01
    m2=0.99
    nc.disp=1
    spn.disp=0
    t=1
    while(nc.disp!=spn.disp & t<10)
    {
      comm.disp=dispersal.ZSM.sloan(op=op.disp,Nk=Nk.disp,sampname=sampname,
                                    spnumi=spnumi.disp,meta.ab=meta.disp,
                                    m1=m1,m2=m2,distinct=TRUE,fix.rich = TRUE)
      (nc.disp=ncol(comm.disp))
      
      comm.drift=drift.ZSM.sloan(op=op.drift,Nk=Nk.drift,sampname=sampname,
                                 spnumi=spnumi.drift,meta.ab=meta.drift,
                                 m1=0.5,m2=0.01,FH=0.95,FL=0.05,sig2.W=4)
      (nc.drift=ncol(comm.drift))
      
      comm=cbind(comm.disp,comm.drift)
      spnuma=ncol(comm)
      spnamea=sample(tree$tip.label,spnuma)
      spc=iCAMP::match.name(name.check = spnamea,both.list = list(pd=pd),tree.list = list(tree=tree),silent = TRUE)
      tree.use=spc$tree
      pd.use=spc$pd
      
      source(paste0(code.wd,"/taxa.bin.phy.r"))
      taxbin=taxa.bin.phy(tree.use,pd.use,d.cut=d.cut,bin.size.limit = bin.size,nworker = nworker,code.wd = code.wd)
      save(taxbin,file=paste0(save.wd,"/" prefixi,".taxabin.rda"))
      tab.sp=table(taxbin$sp.bin$bin.id.new)
      
      bin.num.m=min(ceiling((a.disp+0.1)*length(tab.sp)),length(tab.sp))
      combx<-tab.spx<-list()
      for(x in 1:bin.num.m)
      {
        nnx=choose(length(tab.sp),x)
        if(nnx<1000)
        {
          combx[[x]]=combn(length(tab.sp),x)
          tab.spx[[x]]=sapply(1:ncol(combx[[x]]),function(i){sum(tab.sp[combx[[x]][,i]])})
        }else{
          combx[[x]]=sapply(1:1000,function(xx){sample(length(tab.sp),x)})
          tab.spx[[x]]=sapply(1:ncol(combx[[x]]),function(i){sum(tab.sp[combx[[x]][,i]])})
        }
        names(tab.spx[[x]])=paste0("x",x,"_",1:length(tab.spx[[x]]))
      }
      tabspa=unlist(tab.spx)
      (tab.disp=tabspa[which.min(abs(tabspa-nc.disp))])
      spn.disp=tab.disp[[1]]
      spn.drift=spnuma-spn.disp
      
      k=1
      while(nc.disp!=spn.disp & k<100)
      {
        comm.disp=dispersal.ZSM.sloan(op=op.disp,Nk=Nk.disp,sampname=sampname,
                                      spnumi=spnumi.disp,meta.ab=meta.disp,
                                      m1=m1,m2=m2,distinct=TRUE,fix.rich = TRUE)
        nc.disp=ncol(comm.disp)
        message("Now nc.disp.delta=",nc.disp-spn.disp," t=",t," k=",k,". ",date())
        k=k+1
      }
      message("====Now nc.disp.delta=",nc.disp-spn.disp," t=",t,". ",date())
      t=t+1
    }
    if(nc.disp!=spn.disp){stop("nc.disp!=spn.disp failed")}
    
    tablet=as.numeric(substr(names(tab.disp)[1],2,regexpr("_",names(tab.disp)[1])-1))
    bin.sel=names(tab.sp)[combx[[tablet]][,as.numeric(substring(names(tab.disp)[1],regexpr("_",names(tab.disp)[1])+1))]]
    
    spid.disp=which(taxbin$sp.bin$bin.id.new %in% bin.sel)
    colnames(comm.disp)=rownames(taxbin$sp.bin)[spid.disp]
    
    k=1
    while(nc.drift!=spn.drift & k<400)
    {
      comm.drift=drift.ZSM.sloan(op=op.drift,Nk=Nk.drift,sampname=sampname,
                                 spnumi=spnumi.drift,meta.ab=meta.drift,
                                 m1=0.5,m2=0.01,FH=0.95,FL=0.05,sig2.W=4)
      nc.drift=ncol(comm.drift)
      message("Now nc.drift.delta=",nc.drift-spn.drift," k=",k,". ",date())
      k=k+1
    }
    if(nc.drift!=spn.drift){stop("nc.drift!=spn.drift failed")}
    
    colnames(comm.drift)=rownames(taxbin$sp.bin)[-spid.disp]
    comm=cbind(comm.disp,comm.drift)
    rowSums(comm.disp);rowSums(comm.drift);rowSums(comm)
    
    ABP=ABPcount(comm,disp.spname = colnames(comm.disp),
                 drift.spname = colnames(comm.drift),
                 sp.bin = taxbin$sp.bin,binid.col = 3)
    save.sim(prefixi,comm,pd,tree,save.wd,ABP)
  }
  
  try.fun<-function(fun,...)
  {
    tt=try(fun(...))
    t=1
    while(class(tt)=="try-error" & t<=6)
    {
      tt=try(fun(...))
      t=t+1
    }
    tt
  }
  
  a.disp=0.25 #0.5 #0.25
  a.drift=0.75 #0.5 #0.75
  prefixi=paste0(prefix,".S4.Disp25Drift75")
  try.fun(s46.sim,a.disp=a.disp,a.drift=a.drift,prefixi=prefixi)
  
  a.disp=0.5
  a.drift=0.5
  prefixi=paste0(prefix,".S5.Disp50Drift50")
  #s46.sim(a.disp=a.disp,a.drift=a.drift,prefixi=prefixi)
  try.fun(s46.sim,a.disp=a.disp,a.drift=a.drift,prefixi=prefixi)
  
  a.disp=0.75
  a.drift=0.25
  prefixi=paste0(prefix,".S6.Disp75Drift25")
  try.fun(s46.sim,a.disp=a.disp,a.drift=a.drift,prefixi=prefixi)
  
```

```{r}

# 6 # Scenario 7-9: some species under selection, some under drift
  #####################################
 
  s79.sim<-function(a.sel,a.drift,prefixi,...)
  {
    id.drift=sample(1:length(op),round(length(op)*a.drift))
    meta.drift=MZSM.meta[id.drift]
    op.drift=rep(1,length(id.drift));names(op.drift)=paste0("drift",1:length(id.drift))
    Nk.drift=Nk*a.drift
    spnumi.drift=spnumi*a.drift
    
    nc.drift=1
    spn.drift=0
    t=1
    while(nc.drift!=spn.drift & t<10)
    {
      comm.drift=drift.ZSM.sloan(op=op.drift,Nk=Nk.drift,sampname=sampname,
                                 spnumi=spnumi.drift,meta.ab=meta.drift,
                                 m1=0.5,m2=0.01,FH=0.95,FL=0.05,sig2.W=4)
      (nc.drift=ncol(comm.drift))
      
      
      spnuma=nc.drift+(length(op)-length(op.drift))
      spnamea=sample(tree$tip.label,spnuma)
      spc=iCAMP::match.name(name.check = spnamea,both.list = list(pd=pd),tree.list = list(tree=tree),silent = TRUE)
      tree.use=spc$tree
      pd.use=spc$pd
      source(paste0(code.wd,"/taxa.bin.phy.r"))
      taxbin=taxa.bin.phy(tree.use,pd.use,d.cut=d.cut,bin.size.limit = bin.size,nworker = nworker,code.wd = code.wd)
      save(taxbin,file=paste0(save.wd,"/" prefixi,".taxabin.rda"))))
      tab.sp=table(taxbin$sp.bin$bin.id.new)
      
      bin.num.m=min(ceiling((a.drift+0.1)*length(tab.sp)),length(tab.sp))
      combx<-tab.spx<-list()
      for(x in 1:bin.num.m)
      {
        nnx=choose(length(tab.sp),x)
        if(nnx<1000)
        {
          combx[[x]]=combn(length(tab.sp),x)
          tab.spx[[x]]=sapply(1:ncol(combx[[x]]),function(i){sum(tab.sp[combx[[x]][,i]])})
        }else{
          combx[[x]]=sapply(1:1000,function(xx){sample(length(tab.sp),x)})
          tab.spx[[x]]=sapply(1:ncol(combx[[x]]),function(i){sum(tab.sp[combx[[x]][,i]])})
        }
        names(tab.spx[[x]])=paste0("x",x,"_",1:length(tab.spx[[x]]))
      }
      tabspa=unlist(tab.spx)
      
      (tab.drift=tabspa[which.min(abs(tabspa-nc.drift))])
      spn.drift=tab.drift[[1]]
      k=1
      while(nc.drift!=spn.drift & k<400)
      {
        message("Now nc.drift.delta=",nc.drift-spn.drift," k=",k,". ",date())
        comm.drift=drift.ZSM.sloan(op=op.drift,Nk=Nk.drift,sampname=sampname,
                                   spnumi=spnumi.drift,meta.ab=meta.drift,
                                   m1=0.5,m2=0.01,FH=0.95,FL=0.05,sig2.W=4)
        nc.drift=ncol(comm.drift)
        k=k+1
      }
      message("====Now nc.drift.delta=",nc.drift-spn.drift," t=",t,". ",date())
      t=t+1
    }
    if(nc.drift!=spn.drift){stop("nc.drift!=spn.drift failed.")}
    
    tablet=as.numeric(substr(names(tab.drift)[1],2,regexpr("_",names(tab.drift)[1])-1))
    bin.sel=names(tab.sp)[combx[[tablet]][,as.numeric(substring(names(tab.drift)[1],regexpr("_",names(tab.drift)[1])+1))]]
    
    spid.drift=which(taxbin$sp.bin$bin.id.new %in% bin.sel)
    colnames(comm.drift)=rownames(taxbin$sp.bin)[spid.drift]
    
    spname.sel=rownames(taxbin$sp.bin)[-spid.drift]
    op.sel=op[match(spname.sel, names(op))]
    Nk.sel=Nk*a.sel
    
    sig2.L=0.015
    sig2.H=0.015
    comm.sel=sim_select(op=op.sel,Nk=Nk.sel,sampname=sampname,
                    code.wd=code.wd,FL=0.05,FH=0.95,sig2.L=sig2.L,sig2.H=sig2.H)
    comm.sel=comm.sel[,colSums(comm.sel)>0]
    #if(ncol(comm.sel)!=length(spname.sel)){stop("sel species number not match")}else{message("fine")}
    
    comm=cbind(comm.sel,comm.drift)
    dim(comm)
    spnuma
    
    t=1
    while((ncol(comm)!=spnuma)& t<10)
    {
      spnamea=c(colnames(comm.drift),colnames(comm.sel))
      spc=iCAMP::match.name(name.check = spnamea,both.list = list(pd=pd),tree.list = list(tree=tree),silent = TRUE)
      tree.use=spc$tree
      pd.use=spc$pd
      source(paste0(code.wd,"/taxa.bin.phy.r"))
      taxbin=taxa.bin.phy(tree.use,pd.use,d.cut=d.cut,bin.size.limit = bin.size,nworker = nworker,code.wd = code.wd)
      save(taxbin,file=paste0(save.wd,"/" prefixi,".taxabin.rda"))))
      tab.sp=table(taxbin$sp.bin$bin.id.new)
      
      bin.num.m=min(ceiling((a.drift+0.1)*length(tab.sp)),length(tab.sp))
      combx<-tab.spx<-list()
      for(x in 1:bin.num.m)
      {
        nnx=choose(length(tab.sp),x)
        if(nnx<1000)
        {
          combx[[x]]=combn(length(tab.sp),x)
          tab.spx[[x]]=sapply(1:ncol(combx[[x]]),function(i){sum(tab.sp[combx[[x]][,i]])})
        }else{
          combx[[x]]=sapply(1:1000,function(xx){sample(length(tab.sp),x)})
          tab.spx[[x]]=sapply(1:ncol(combx[[x]]),function(i){sum(tab.sp[combx[[x]][,i]])})
        }
        names(tab.spx[[x]])=paste0("x",x,"_",1:length(tab.spx[[x]]))
      }
      tabspa=unlist(tab.spx)
      (tab.drift=tabspa[which.min(abs(tabspa-nc.drift))])
      spn.drift=tab.drift[[1]]
      k=1
      while(nc.drift!=spn.drift & k<400)
      {
        comm.drift=drift.ZSM.sloan(op=op.drift,Nk=Nk.drift,sampname=sampname,
                                   spnumi=spnumi.drift,meta.ab=meta.drift,
                                   m1=0.5,m2=0.01,FH=0.95,FL=0.05,sig2.W=4)
        nc.drift=ncol(comm.drift)
        message("Now nc.drift.delta=",nc.drift-spn.drift," k=",k,". ",date())
        k=k+1
      }
      if(nc.drift==spn.drift)
      {
        tablet=as.numeric(substr(names(tab.drift)[1],2,regexpr("_",names(tab.drift)[1])-1))
        bin.sel=names(tab.sp)[combx[[tablet]][,as.numeric(substring(names(tab.drift)[1],regexpr("_",names(tab.drift)[1])+1))]]
        
        spid.drift=which(taxbin$sp.bin$bin.id.new %in% bin.sel)
        colnames(comm.drift)=rownames(taxbin$sp.bin)[spid.drift]
        
        spname.sel=rownames(taxbin$sp.bin)[-spid.drift]
        op.sel=op[match(spname.sel, names(op))]
        Nk.sel=Nk*a.sel
        
        sig2.L=0.015
        sig2.H=0.015
        comm.sel=sim_select(op=op.sel,Nk=Nk.sel,sampname=sampname,
                        code.wd=code.wd,FL=0.05,FH=0.95,sig2.L=sig2.L,sig2.H=sig2.H)
        comm.sel=comm.sel[,colSums(comm.sel)>0]
        #if(ncol(comm.sel)!=length(spname.sel)){message("sel species number not match")}else{message("fine")}
        
        comm=cbind(comm.sel,comm.drift)  
        spnuma=length(spnamea)
        print(dim(comm))
        print(spnuma)
      }
      message("-----t=",t,". dert=",ncol(comm)-spnuma,". ",date())
      t=t+1
    }
    if(ncol(comm)!=spnuma){stop("(ncol(comm)!=spnuma) failed")}
    ABP=ABPcount(comm,sel.spname = colnames(comm.sel),
                 drift.spname = colnames(comm.drift),
                 sp.bin = taxbin$sp.bin,binid.col = 3)
    save.sim(prefixi,comm,pd,tree,save.wd,ABP)
  }
  
  a.sel=0.25
  a.drift=0.75
  prefixi=paste0(prefix,".S7.Sel25Drift75")
  try.fun(s79.sim,a.sel=a.sel,a.drift=a.drift,prefixi=prefixi)
  
  a.sel=0.5
  a.drift=0.5
  prefixi=paste0(prefix,".S8.Sel50Drift50")
  try.fun(s79.sim,a.sel=a.sel,a.drift=a.drift,prefixi=prefixi)
  
  a.sel=0.75
  a.drift=0.25
  prefixi=paste0(prefix,".S9.Sel75Drift25")
  try.fun(s79.sim,a.sel=a.sel,a.drift=a.drift,prefixi=prefixi)
  
```

```{r}

# 7 # Scenario 10-12: some species under selection, some under extreme dispersal
  #####################################
  
  s1012.sim<-function(a.sel,a.disp,prefixi,...)
  {
    sig2.L=0.015
    sig2.H=0.015
    id.disp=sample(1:length(op),round(length(op)*a.disp))
    meta.disp=MZSM.meta[id.disp]
    op.disp=rep(1,length(id.disp));names(op.disp)=paste0("disp",1:length(id.disp))
    Nk.disp=Nk*a.disp
    spnumi.disp=spnumi*a.disp
    
    m1=0.01
    m2=0.99
    
    nc.disp=1
    spn.disp=0
    t=1
    while(nc.disp!=spn.disp & t<10)
    {
      comm.disp=dispersal.ZSM.sloan(op=op.disp,Nk=Nk.disp,sampname=sampname,
                                    spnumi=spnumi.disp,meta.ab=meta.disp,
                                    m1=m1,m2=m2,distinct=TRUE,fix.rich = TRUE)
      (nc.disp=ncol(comm.disp))
      
      
      spnuma=nc.disp+(length(op)-length(op.disp))
      spnamea=sample(tree$tip.label,spnuma)
      spc=iCAMP::match.name(name.check = spnamea,both.list = list(pd=pd),tree.list = list(tree=tree),silent = TRUE)
      tree.use=spc$tree
      pd.use=spc$pd
      source(paste0(code.wd,"/taxa.bin.phy.r"))
      taxbin=taxa.bin.phy(tree.use,pd.use,d.cut=d.cut,bin.size.limit = bin.size,nworker = nworker,code.wd = code.wd)
      save(taxbin,file=paste0(save.wd,"/" prefixi,".taxabin.rda"))))
      tab.sp=table(taxbin$sp.bin$bin.id.new)
      
      bin.num.m=min(ceiling((a.disp+0.1)*length(tab.sp)),length(tab.sp))
      combx<-tab.spx<-list()
      for(x in 1:bin.num.m)
      {
        nnx=choose(length(tab.sp),x)
        if(nnx<1000)
        {
          combx[[x]]=combn(length(tab.sp),x)
          tab.spx[[x]]=sapply(1:ncol(combx[[x]]),function(i){sum(tab.sp[combx[[x]][,i]])})
        }else{
          combx[[x]]=sapply(1:1000,function(xx){sample(length(tab.sp),x)})
          tab.spx[[x]]=sapply(1:ncol(combx[[x]]),function(i){sum(tab.sp[combx[[x]][,i]])})
        }
        names(tab.spx[[x]])=paste0("x",x,"_",1:length(tab.spx[[x]]))
      }
      tabspa=unlist(tab.spx)
      
      (tab.disp=tabspa[which.min(abs(tabspa-nc.disp))])
      spn.disp=tab.disp[[1]]
      k=1
      while(nc.disp!=spn.disp & k<400)
      {
        comm.disp=dispersal.ZSM.sloan(op=op.disp,Nk=Nk.disp,sampname=sampname,
                                      spnumi=spnumi.disp,meta.ab=meta.disp,
                                      m1=m1,m2=m2,distinct=TRUE,fix.rich = TRUE)
        nc.disp=ncol(comm.disp)
        message("Now nc.disp.delta=",nc.disp-spn.disp," k=",k,". ",date())
        k=k+1
      }
      message("====Now nc.disp.delta=",nc.disp-spn.disp," t=",t,". ",date())
      t=t+1
    }
    if(nc.disp!=spn.disp){stop("nc.disp!=spn.disp failed")}
    
    tablet=as.numeric(substr(names(tab.disp)[1],2,regexpr("_",names(tab.disp)[1])-1))
    bin.sel=names(tab.sp)[combx[[tablet]][,as.numeric(substring(names(tab.disp)[1],regexpr("_",names(tab.disp)[1])+1))]]
    
    spid.disp=which(taxbin$sp.bin$bin.id.new %in% bin.sel)
    colnames(comm.disp)=rownames(taxbin$sp.bin)[spid.disp]
    
    spname.sel=rownames(taxbin$sp.bin)[-spid.disp]
    op.sel=op[match(spname.sel, names(op))]
    Nk.sel=Nk*a.sel
    
    comm.sel=sim_select(op=op.sel,Nk=Nk.sel,sampname=sampname,
                    code.wd=code.wd,FL=0.05,FH=0.95,sig2.L=sig2.L,sig2.H=sig2.H)
    comm.sel=comm.sel[,colSums(comm.sel)>0]
    #if(ncol(comm.sel)!=length(spname.sel)){stop("sel species number not match")}else{message("fine")}
    
    comm=cbind(comm.sel,comm.disp)
    dim(comm)
    spnuma
    t=1
    while((ncol(comm)!=spnuma) & t<10)
    {
      spnamea=c(colnames(comm.disp),colnames(comm.sel))
      spc=iCAMP::match.name(name.check = spnamea,both.list = list(pd=pd),tree.list = list(tree=tree),silent = TRUE)
      tree.use=spc$tree
      pd.use=spc$pd
      source(paste0(code.wd,"/taxa.bin.phy.r"))
      taxbin=taxa.bin.phy(tree.use,pd.use,d.cut=d.cut,bin.size.limit = bin.size,nworker = nworker,code.wd = code.wd)
      save(taxbin,file=paste0(save.wd,"/" prefixi,".taxabin.rda"))))
      tab.sp=table(taxbin$sp.bin$bin.id.new)
      
      bin.num.m=min(ceiling((a.disp+0.1)*length(tab.sp)),length(tab.sp))
      combx<-tab.spx<-list()
      for(x in 1:bin.num.m)
      {
        nnx=choose(length(tab.sp),x)
        if(nnx<1000)
        {
          combx[[x]]=combn(length(tab.sp),x)
          tab.spx[[x]]=sapply(1:ncol(combx[[x]]),function(i){sum(tab.sp[combx[[x]][,i]])})
        }else{
          combx[[x]]=sapply(1:1000,function(xx){sample(length(tab.sp),x)})
          tab.spx[[x]]=sapply(1:ncol(combx[[x]]),function(i){sum(tab.sp[combx[[x]][,i]])})
        }
        names(tab.spx[[x]])=paste0("x",x,"_",1:length(tab.spx[[x]]))
      }
      tabspa=unlist(tab.spx)
      
      (tab.disp=tabspa[which.min(abs(tabspa-nc.disp))])
      
      spn.disp=tab.disp[[1]]
      k=1
      while(nc.disp!=spn.disp & k<400)
      {
        comm.disp=dispersal.ZSM.sloan(op=op.disp,Nk=Nk.disp,sampname=sampname,
                                      spnumi=spnumi.disp,meta.ab=meta.disp,
                                      m1=m1,m2=m2,distinct=TRUE,fix.rich = TRUE)
        nc.disp=ncol(comm.disp)
        message("Now nc.disp.delta=",nc.disp-spn.disp," k=",k,". ",date())
        k=k+1
      }
      if(nc.disp==spn.disp)
      {
        tablet=as.numeric(substr(names(tab.disp)[1],2,regexpr("_",names(tab.disp)[1])-1))
        bin.sel=names(tab.sp)[combx[[tablet]][,as.numeric(substring(names(tab.disp)[1],regexpr("_",names(tab.disp)[1])+1))]]
        
        spid.disp=which(taxbin$sp.bin$bin.id.new %in% bin.sel)
        colnames(comm.disp)=rownames(taxbin$sp.bin)[spid.disp]
        
        spname.sel=rownames(taxbin$sp.bin)[-spid.disp]
        op.sel=op[match(spname.sel, names(op))]
        Nk.sel=Nk*a.sel
        
        sig2.L=0.015
        sig2.H=0.015
        comm.sel=sim_select(op=op.sel,Nk=Nk.sel,sampname=sampname,
                        code.wd=code.wd,FL=0.05,FH=0.95,sig2.L=sig2.L,sig2.H=sig2.H)
        comm.sel=comm.sel[,colSums(comm.sel)>0]
        #if(ncol(comm.sel)!=length(spname.sel)){message("sel species number not match")}else{message("fine")}
        
        comm=cbind(comm.sel,comm.disp)
        spnuma=length(spnamea)
        print(dim(comm))
        print(spnuma)
      }
      message("-----t=",t,". dert=",ncol(comm)-spnuma,". ",date())
      t=t+1
    }
    if(ncol(comm)!=spnuma){stop("(ncol(comm)!=spnuma) failed.")}
    ABP=ABPcount(comm,sel.spname = colnames(comm.sel),
                 disp.spname = colnames(comm.disp),
                 sp.bin = taxbin$sp.bin,binid.col = 3)
    save.sim(prefixi,comm,pd,tree,save.wd,ABP)
  }
  
  a.sel=0.25
  a.disp=0.75
  prefixi=paste0(prefix,".S10.Sel25Disp75")
  try.fun(s1012.sim,a.sel=a.sel,a.disp=a.disp,prefixi=prefixi)
  
  a.sel=0.5
  a.disp=0.5
  prefixi=paste0(prefix,".S11.Sel50Disp50")
  try.fun(s1012.sim,a.sel=a.sel,a.disp=a.disp,prefixi=prefixi)
  
  a.sel=0.75
  a.disp=0.25
  prefixi=paste0(prefix,".S12.Sel75Disp25")
  try.fun(s1012.sim,a.sel=a.sel,a.disp=a.disp,prefixi=prefixi)
  
  #####################################
  # 8 # Scenario 13-15: some species under selection, some under extreme dispersal, and others under drift
  #####################################
  
  s1315.sim<-function(a.sel,a.disp,a.drift,prefixi,...)
  {
    id.disp=sample(1:length(op),round(length(op)*a.disp))
    meta.disp=MZSM.meta[id.disp]
    op.disp=rep(1,length(id.disp));names(op.disp)=paste0("disp",1:length(id.disp))
    Nk.disp=Nk*a.disp
    spnumi.disp=spnumi*a.disp
    
    id.drift=sample((1:length(op))[-id.disp],round(length(op)*a.drift))
    meta.drift=MZSM.meta[id.drift]
    op.drift=rep(1,length(id.drift));names(op.drift)=paste0("drift",1:length(id.drift))
    Nk.drift=Nk*a.drift
    spnumi.drift=spnumi*a.drift
    
    
    m1.disp=0.01
    m2.disp=0.99
    comm.disp=dispersal.ZSM.sloan(op=op.disp,Nk=Nk.disp,sampname=sampname,
                                  spnumi=spnumi.disp,meta.ab=meta.disp,
                                  m1=m1.disp,m2=m2.disp,distinct=TRUE,fix.rich = TRUE)
    (nc.disp=ncol(comm.disp))
    
    m1.drift=0.5
    m2.drift=0.01
    comm.drift=drift.ZSM.sloan(op=op.drift,Nk=Nk.drift,sampname=sampname,
                               spnumi=spnumi.drift,meta.ab=meta.drift,
                               m1=m1.drift,m2=m2.drift,FH=0.95,FL=0.05,sig2.W=4)
    (nc.drift=ncol(comm.drift))
    
    spnuma=nc.disp+nc.drift+(length(op)-length(op.disp)-length(op.drift))
    spnamea=sample(tree$tip.label,spnuma)
    spc=iCAMP::match.name(name.check = spnamea,both.list = list(pd=pd),tree.list = list(tree=tree),silent = TRUE)
    tree.use=spc$tree
    pd.use=spc$pd
    source(paste0(code.wd,"/taxa.bin.phy.r"))
    taxbin=taxa.bin.phy(tree.use,pd.use,d.cut=d.cut,bin.size.limit = bin.size,nworker = nworker,code.wd = code.wd)
    save(taxbin,file=paste0(save.wd,"/" prefixi,".taxabin.rda"))))
    tab.sp=table(taxbin$sp.bin$bin.id.new)
    
    bin.num.m=min(ceiling((a.disp+0.1)*length(tab.sp)),length(tab.sp))
    combx<-tab.spx<-list()
    for(x in 1:bin.num.m)
    {
      nnx=choose(length(tab.sp),x)
      if(nnx<1000)
      {
        combx[[x]]=combn(length(tab.sp),x)
        tab.spx[[x]]=sapply(1:ncol(combx[[x]]),function(i){sum(tab.sp[combx[[x]][,i]])})
      }else{
        combx[[x]]=sapply(1:1000,function(xx){sample(length(tab.sp),x)})
        tab.spx[[x]]=sapply(1:ncol(combx[[x]]),function(i){sum(tab.sp[combx[[x]][,i]])})
      }
      names(tab.spx[[x]])=paste0("x",x,"_",1:length(tab.spx[[x]]))
    }
    tabspa=unlist(tab.spx)
    
    (tab.disp=tabspa[which.min(abs(tabspa-nc.disp))])
    
    spn.disp=tab.disp[[1]]
    k=1
    while(nc.disp!=spn.disp & k<400)
    {
      comm.disp=dispersal.ZSM.sloan(op=op.disp,Nk=Nk.disp,sampname=sampname,
                                    spnumi=spnumi.disp,meta.ab=meta.disp,
                                    m1=m1.disp,m2=m2.disp,distinct=TRUE,fix.rich = TRUE)
      nc.disp=ncol(comm.disp)
      message("Now nc.disp.delta=",nc.disp-spn.disp," k=",k,". ",date())
      k=k+1
    }
    
    tablet=as.numeric(substr(names(tab.disp)[1],2,regexpr("_",names(tab.disp)[1])-1))
    bin.sel=names(tab.sp)[combx[[tablet]][,as.numeric(substring(names(tab.disp)[1],regexpr("_",names(tab.disp)[1])+1))]]
    
    spid.disp=which(taxbin$sp.bin$bin.id.new %in% bin.sel)
    colnames(comm.disp)=rownames(taxbin$sp.bin)[spid.disp]
    
    tab.sp.old=tab.sp
    tab.sp=tab.sp.old[!(names(tab.sp.old) %in% bin.sel)]
    
    bin.num.m=min(ceiling((a.drift+0.1)*length(tab.sp)),length(tab.sp))
    combx<-tab.spx<-list()
    for(x in 1:bin.num.m)
    {
      nnx=choose(length(tab.sp),x)
      if(nnx<1000)
      {
        combx[[x]]=combn(length(tab.sp),x)
        tab.spx[[x]]=sapply(1:ncol(combx[[x]]),function(i){sum(tab.sp[combx[[x]][,i]])})
      }else{
        combx[[x]]=sapply(1:1000,function(xx){sample(length(tab.sp),x)})
        tab.spx[[x]]=sapply(1:ncol(combx[[x]]),function(i){sum(tab.sp[combx[[x]][,i]])})
      }
      names(tab.spx[[x]])=paste0("x",x,"_",1:length(tab.spx[[x]]))
    }
    tabspa=unlist(tab.spx)
    
    (tab.drift=tabspa[which.min(abs(tabspa-nc.drift))])
    
    spn.drift=tab.drift[[1]]
    k=1
    while(nc.drift!=spn.drift & k<400)
    {
      comm.drift=drift.ZSM.sloan(op=op.drift,Nk=Nk.drift,sampname=sampname,
                                 spnumi=spnumi.drift,meta.ab=meta.drift,
                                 m1=m1.drift,m2=m2.drift,FH=0.95,FL=0.05,sig2.W=4)
      nc.drift=ncol(comm.drift)
      message("Now nc.drift.delta=",nc.drift-spn.drift," k=",k,". ",date())
      k=k+1
    }
    if(nc.drift!=spn.drift){stop("nc.drift!=spn.drift failed.")}
    
    tablet=as.numeric(substr(names(tab.drift)[1],2,regexpr("_",names(tab.drift)[1])-1))
    bin.sel=names(tab.sp)[combx[[tablet]][,as.numeric(substring(names(tab.drift)[1],regexpr("_",names(tab.drift)[1])+1))]]
    
    spid.drift=which(taxbin$sp.bin$bin.id.new %in% bin.sel)
    colnames(comm.drift)=rownames(taxbin$sp.bin)[spid.drift]
    
    
    spname.sel=rownames(taxbin$sp.bin)[-(c(spid.drift,spid.disp))]
    op.sel=op[match(spname.sel, names(op))]
    Nk.sel=Nk*a.sel
    
    sig2.L=0.015
    sig2.H=0.015
    comm.sel=sim_select(op=op.sel,Nk=Nk.sel,sampname=sampname,
                    code.wd=code.wd,FL=0.05,FH=0.95,sig2.L=sig2.L,sig2.H=sig2.H)
    comm.sel=comm.sel[,colSums(comm.sel)>0]
    #if(ncol(comm.sel)!=length(spname.sel)){stop("sel species number not match")}else{message("fine")}
    
    comm=cbind(comm.sel,comm.disp,comm.drift)
    dim(comm)
    spnuma
    
    t=1
    while((ncol(comm)!=spnuma)& t<10)
    {
      spnamea=c(colnames(comm.disp),colnames(comm.drift),colnames(comm.sel))
      spc=iCAMP::match.name(name.check = spnamea,both.list = list(pd=pd),tree.list = list(tree=tree),silent = TRUE)
      tree.use=spc$tree
      pd.use=spc$pd
      source(paste0(code.wd,"/taxa.bin.phy.r"))
      taxbin=taxa.bin.phy(tree.use,pd.use,d.cut=d.cut,bin.size.limit = bin.size,nworker = nworker,code.wd = code.wd)
      save(taxbin,file=paste0(save.wd,"/" prefixi,".taxabin.rda"))))
      tab.sp=table(taxbin$sp.bin$bin.id.new)
      
      
      bin.num.m=min(ceiling((a.disp+0.1)*length(tab.sp)),length(tab.sp))
      combx<-tab.spx<-list()
      for(x in 1:bin.num.m)
      {
        nnx=choose(length(tab.sp),x)
        if(nnx<1000)
        {
          combx[[x]]=combn(length(tab.sp),x)
          tab.spx[[x]]=sapply(1:ncol(combx[[x]]),function(i){sum(tab.sp[combx[[x]][,i]])})
        }else{
          combx[[x]]=sapply(1:1000,function(xx){sample(length(tab.sp),x)})
          tab.spx[[x]]=sapply(1:ncol(combx[[x]]),function(i){sum(tab.sp[combx[[x]][,i]])})
        }
        names(tab.spx[[x]])=paste0("x",x,"_",1:length(tab.spx[[x]]))
      }
      tabspa=unlist(tab.spx)
      (tab.disp=tabspa[which.min(abs(tabspa-nc.disp))])
      
      spn.disp=tab.disp[[1]]
      k=1
      while(nc.disp!=spn.disp & k<400)
      {
        comm.disp=dispersal.ZSM.sloan(op=op.disp,Nk=Nk.disp,sampname=sampname,
                                      spnumi=spnumi.disp,meta.ab=meta.disp,
                                      m1=m1.disp,m2=m2.disp,distinct=TRUE,fix.rich = TRUE)
        nc.disp=ncol(comm.disp)
        message("Now nc.disp.delta=",nc.disp-spn.disp," k=",k,". ",date())
        k=k+1
      }
      if(nc.disp==spn.disp)
      {
        tablet=as.numeric(substr(names(tab.disp)[1],2,regexpr("_",names(tab.disp)[1])-1))
        bin.sel=names(tab.sp)[combx[[tablet]][,as.numeric(substring(names(tab.disp)[1],regexpr("_",names(tab.disp)[1])+1))]]
        
        spid.disp=which(taxbin$sp.bin$bin.id.new %in% bin.sel)
        colnames(comm.disp)=rownames(taxbin$sp.bin)[spid.disp]
        
        tab.sp.old=tab.sp
        tab.sp=tab.sp.old[!(names(tab.sp.old) %in% bin.sel)]
        
        
        bin.num.m=min(ceiling((a.drift+0.1)*length(tab.sp)),length(tab.sp))
        combx<-tab.spx<-list()
        for(x in 1:bin.num.m)
        {
          nnx=choose(length(tab.sp),x)
          if(nnx<1000)
          {
            combx[[x]]=combn(length(tab.sp),x)
            tab.spx[[x]]=sapply(1:ncol(combx[[x]]),function(i){sum(tab.sp[combx[[x]][,i]])})
          }else{
            combx[[x]]=sapply(1:1000,function(xx){sample(length(tab.sp),x)})
            tab.spx[[x]]=sapply(1:ncol(combx[[x]]),function(i){sum(tab.sp[combx[[x]][,i]])})
          }
          names(tab.spx[[x]])=paste0("x",x,"_",1:length(tab.spx[[x]]))
        }
        tabspa=unlist(tab.spx)
        
        (tab.drift=tabspa[which.min(abs(tabspa-nc.drift))])
        
        spn.drift=tab.drift[[1]]
        k=1
        while(nc.drift!=spn.drift & k<400)
        {
          message("Now nc.drift.delta=",nc.drift-spn.drift," k=",k,". ",date())
          comm.drift=drift.ZSM.sloan(op=op.drift,Nk=Nk.drift,sampname=sampname,
                                     spnumi=spnumi.drift,meta.ab=meta.drift,
                                     m1=m1.drift,m2=m2.drift,FH=0.95,FL=0.05,sig2.W=4)
          nc.drift=ncol(comm.drift)
          k=k+1
        }
        if(nc.drift==spn.drift)
        {
          tablet=as.numeric(substr(names(tab.drift)[1],2,regexpr("_",names(tab.drift)[1])-1))
          bin.sel=names(tab.sp)[combx[[tablet]][,as.numeric(substring(names(tab.drift)[1],regexpr("_",names(tab.drift)[1])+1))]]
          
          spid.drift=which(taxbin$sp.bin$bin.id.new %in% bin.sel)
          colnames(comm.drift)=rownames(taxbin$sp.bin)[spid.drift]
          
          spname.sel=rownames(taxbin$sp.bin)[-(c(spid.drift,spid.disp))]
          op.sel=op[match(spname.sel, names(op))]
          Nk.sel=Nk*a.sel
          
          sig2.L=0.015
          sig2.H=0.015
          comm.sel=sim_select(op=op.sel,Nk=Nk.sel,sampname=sampname,
                          code.wd=code.wd,FL=0.05,FH=0.95,sig2.L=sig2.L,sig2.H=sig2.H)
          comm.sel=comm.sel[,colSums(comm.sel)>0]
          #if(ncol(comm.sel)!=length(spname.sel)){message("sel species number not match")}else{message("fine")}
          
          comm=cbind(comm.sel,comm.disp,comm.drift)
          spnuma=length(spnamea)
          print(dim(comm))
          print(spnuma)
        }
      }
      message("-----t=",t,". dert=",ncol(comm)-spnuma,". ",date())
      t=t+1
    }
    if(ncol(comm)!=spnuma){stop("ncol(comm)!=spnuma failed.")}
    ABP=ABPcount(comm,sel.spname = colnames(comm.sel),
                 disp.spname = colnames(comm.disp),
                 drift.spname = colnames(comm.drift),
                 sp.bin = taxbin$sp.bin,binid.col = 3)
    save.sim(prefixi,comm,pd,tree,save.wd,ABP)
  }
  
  a.sel=0.25
  a.disp=0.25
  a.drift=0.5
  prefixi=paste0(prefix,".S13.Sel25Disp25Drift50")
  try.fun(s1315.sim,a.sel=a.sel,a.disp=a.disp,a.drift=a.drift,prefixi=prefixi)
  
  a.sel=0.25
  a.disp=0.5
  a.drift=0.25
  prefixi=paste0(prefix,".S14.Sel25Disp50Drift25")
  try.fun(s1315.sim,a.sel=a.sel,a.disp=a.disp,a.drift=a.drift,prefixi=prefixi)
  
  a.sel=0.5
  a.disp=0.25
  a.drift=0.25
  prefixi=paste0(prefix,".S15.Sel50Disp25Drift25")
  try.fun(s1315.sim,a.sel=a.sel,a.disp=a.disp,a.drift=a.drift,prefixi=prefixi)

```

``` {r}

ds=0.2 # we used different ds, including 0.1, 0.2, 0.4.
bszs=c(3,6,12,24,48)

# Low phylogenetic signal scenario # JS
op=ops$JS$openv 
for(bz in 1:length(bszs))
{
  bin.size=bszs[bz]
  prefix=paste0("JSb",bin.size,"d",sub("\\.","",ds))
  simmu(prefix=prefix,op=op,bin.size=bin.size)
}

# Medium phylogenetic signal scenario # BM
op=ops$BM$openv
for(bz in 1:length(bszs))
{
  bin.size=bszs[bz]
  prefix=paste0("BMb",bin.size,"d",sub("\\.","",ds))
  simmu(prefix=prefix,op=op,bin.size=bin.size)
}

# High phylogenetic signal scenario # AD
op=ops$ACDC$openv
for(bz in 1:length(bszs))
{
  bin.size=bszs[bz]
  prefix=paste0("ADb",bin.size,"d02")
  simmu(prefix=prefix,op=op,bin.size=bin.size)
}
